cardinalR: Generating interesting high-dimensional data structures

### 2025-09-20 paper-cardinalR.Rmd
Updated sphere section

### Sphere

Sphere-shaped structures (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere)"} else { "Figure \\@ref(fig:sphere-proj)"})`) are useful for evaluating how dimension reduction and clustering algorithms handle curved, symmetric manifolds in high-dimensional spaces. The package provides functions to generate a variety of spherical forms, including simple circles, uniform spheres, grid-based spheres, and complex arrangements like clustered spheres within a larger sphere. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:sphere-tb-html)', '\\@ref(tab:sphere-tb-pdf)'))` summarizes these functions.

@@

#### `gen_circle()`

The function `gen_circle(n, p)` generates a $p$-dimensional dataset of $n$ observations, where the first two dimensions form a unit circle, and the remaining dimensions are structured sinusoidal extensions of the angular parameter with progressively smaller scale (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) a"} else { "Figure \\@ref(fig:sphere-proj) a"})`).

A latent angle variable $\theta$ is uniformly sampled from the interval $[0, 2\pi]$. Coordinates in the first two dimensions represent a perfect circle on the plane: $X_1 = \cos(\theta), \quad X_2 = \sin(\theta)$. For dimensions $X_3$ through $X_p$, sinusoidal transformations of the angle $\theta$ are introduced. The first component is a scaling factor that decreases with the dimension index, defined as $\text{scale}_j = \sqrt{(0.5)^{j-2}}$ for $j = 3, \dots, p$. The second component is a phase shift that is proportional to the dimension index, specifically designed to decorrelate the curves, given by the formula $\phi_j = (j - 2)\pi/2p$. Each additional dimension is computed as: $X_j = \text{scale}_{j}\sin(\theta + \phi_j), \quad j = 3, \dots, p$.

@@

#### `gen_curvycycle()`

The `gen_curvycycle(n, p)` function generates a $p$-dimensional dataset of $n$ observations lying on a curved closed loop with controlled high-dimensional sinusoidal deviations (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) b"} else { "Figure \\@ref(fig:sphere-proj) b"})`). The structure forms a one-dimensional nonlinear cycle embedded in higher-dimensional space.

A latent angle variable $\theta$ is uniformly sampled from the interval $[0, 2\pi]$. The first three dimensions define a non-circular closed curve, referred to as a "curvy cycle". In this configuration, $X_1 = \cos(\theta)$ represents horizontal oscillation, while $X_2 = \sqrt{3}/3 + \sin(\theta)$ introduces a vertical offset to avoid centering the curve at the origin. Additionally, $X_3 = 1/3\cos(3\theta)$ introduces a third harmonic perturbation that intricately folds the curve three times along its path, creating a unique and complex shape that oscillates in both dimensions while incorporating the effects of the harmonic perturbation.

@@

#### `gen_unifsphere()`

The `gen_unifsphere(n, p, r)` function generates a $p$-dimensional dataset of $n$ observations distributed approximately uniformly on the surface of a $3\text{-}D$ sphere of radius $r$, with additional Gaussian noise dimensions added when $p > 3$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) c"} else { "Figure \\@ref(fig:sphere-proj) c"})`).

Each observation lies on the surface of a sphere in $3\text{-}D$, constructed by generating $u \sim U(-1, 1)$ which represents the cosine of the polar angle $\phi$ and $\theta \sim U(0, 2\pi)$ which represents the azimuthal angle. The corresponding Cartesian coordinates are calculated as; $X_1 = r\sqrt{1 - u^2}\cos(\theta)$, $X_2 = r\sqrt{1 - u^2}\sin(\theta)$, $X_3 = ru$, which gives points uniformly distributed on the surface of a $3\text{-}D$ sphere (not within). For $p > 3$, additional noise dimensions $X_4$ to $X_p$ are generated.

@@

#### `gen_gridedsphere()`

The `gen_gridedsphere(n, p)` function generates a $p$-dimensional dataset of approximately $n$ points evenly distributed on the surface of a $(p!-!1)$-sphere (the unit sphere embedded in $\mathbb{R}^p$) (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) d"} else { "Figure \\@ref(fig:sphere-proj) d"})`).

The construction is based on a regular grid in spherical coordinates with $(p-1)$ angular variables. Specifically, for dimensions $j = 1, \dots, p-2$, the polar angles are defined on $[0, \pi]$. For the last angle ($j = p-1$), the azimuthal angle is defined on $[0, 2\pi]$.

The number of grid steps along each angular dimension is determined by factoring $n$ into $(p-1)$ approximately equal integers via `gen_nproduct(n, p - 1)`.

Each grid point is then mapped to Cartesian coordinates using the standard hyperspherical-to-Cartesian transformation:

@@

Thus, the output is a deterministic grid of points lying exactly on the unit $(p-1)$-sphere, with no added noise dimensions.

@@

#### `gen_clusteredspheres()`

The `gen_clusteredspheres(n, k, p, r, loc)` function generates a synthetic dataset of $n_1 + kn_2$ observations in $p$-dimensional space, consisting of one large sphere of radius $r_1$ and $k$ smaller spheres of radius $r_2$, each centered at a different random location (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) e"} else { "Figure \\@ref(fig:sphere-proj) e"})`).

A large uniform sphere centered at the origin is created by sampling $n_1$ points uniformly on the surface of a $p\text{-}D$ sphere with a radius of $r_1$. The sampling is executed using the function `gen_unifsphere(n_1, p, r_1)`, which generates the desired points in the specified dimensional space.

In generation of $k$ smaller uniform spheres, each sphere contains $n_2$ points that are sampled uniformly on a sphere with a radius of $r_2$. These spheres are positioned at distinct random locations in $p$-space, with the center of each sphere being drawn from a normal distribution $N(0, \texttt{loc}^2 I_p)$.

Points on spheres are generated using the standard hyperspherical method, which involves sampling $u \sim U(-1, 1)$ to determine the cosine of the polar angle, and sampling $\theta \sim U(0, 2\pi)$ to determine the azimuthal angle (for $3\text{-}D$). Each observation is classified by cluster, with labels such as "big" for the large central sphere and "small_1" to "small_k" for the smaller spheres.

@@

#### `gen_hemisphere()`

The `gen_hemisphere(n, p)` function generates a $p$-dimensional dataset of $n$ observations distributed approximately uniformly on a $4\text{-}D$ hemisphere, optionally extended with Gaussian noise in additional dimensions when $p > 4$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) f"} else { "Figure \\@ref(fig:sphere-proj) f"})`).

Each observation is situated on a restricted $4\text{-}D$ spherical surface, defined by spherical coordinates. The azimuthal angle $\theta_1 \sim U(0, \pi)$ in the $(x_1, x_2)$ plane, while the elevation angle $\theta_2 \sim U(0, \pi)$ in the $(x_2, x_3)$ plane. Additionally, $\theta_3 \sim U(0, \pi/2)$ in the $(x_3, x_4)$ plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into $4\text{-}D$ Cartesian space: $X_1 = \sin(\theta_1)\cos(\theta_2)$, $X_2 = \sin(\theta_1)\sin(\theta_2)$, $X_3 = \cos(\theta_1)\cos(\theta_3)$, $X_4 = \cos(\theta_1)\sin(\theta_3)$. This produces points on one side of a $4\text{-}D$ unit sphere, effectively generating a $4\text{-}D$ hemisphere. For $p > 4$, additional noise dimensions are added to embed the structure in higher dimensions.

>>>>

#### Sphere

Sphere-shaped structures (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere)"} else { "Figure \\@ref(fig:sphere-proj)"})`) are useful for evaluating how dimension reduction and clustering algorithms handle curved, symmetric manifolds in high-dimensional spaces. The functions generate a variety of spherical forms, including simple circles, uniform spheres, grid-based spheres, and complex arrangements like clustered spheres within a larger sphere. The first few coordinates define the main geometric form (circle, cycle, sphere, or hemisphere), while higher-dimensional embeddings are achieved by adding noise dimensions. For the circle and curvy cycle, this occurs when $p > 2$ and $p > 3$, respectively, while for spheres and hemispheres it occurs when $p > 3$ or $p > 4$. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:sphere-tb-html)', '\\@ref(tab:sphere-tb-pdf)'))` summarizes these functions.

@@

The simplest case, `gen_circle(n, p)` creates a unit circle in two dimensions, with the remaining dimensions forming sinusoidal extensions of the angular parameter at progressively smaller scales (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) a"} else { "Figure \\@ref(fig:sphere-proj) a"})`). Let a latent angle variable $\theta$ is uniformly sampled from the interval $[0, 2\pi]$. Coordinates in the first two dimensions represent a perfect circle on the plane: $X_1 = \cos(\theta), \quad X_2 = \sin(\theta)$. For dimensions $X_3$ through $X_p$, sinusoidal transformations of the angle $\theta$ are introduced. The first component is a scaling factor that decreases with the dimension index, defined as $\text{scale}_j = \sqrt{(0.5)^{j-2}}$ for $j = 3, \dots, p$. The second component is a phase shift that is proportional to the dimension index, specifically designed to decorrelate the curves, given by the formula $\phi_j = (j - 2)\pi/2p$. Each additional dimension is computed as: $X_j = \text{scale}_{j}\sin(\theta + \phi_j), \quad j = 3, \dots, p$.

@@

For the one-dimensional nonlinear cycle embedded in $p\text{-}D$ space, `gen_curvycycle(n, p)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) b"} else { "Figure \\@ref(fig:sphere-proj) b"})`), let a latent angle variable $\theta$ is uniformly sampled from the interval $[0, 2\pi]$. The first three dimensions define a non-circular closed curve, referred to as a "curvy cycle". In this configuration, $X_1 = \cos(\theta)$ represents horizontal oscillation, while $X_2 = \sqrt{3}/3 + \sin(\theta)$ introduces a vertical offset to avoid centering the curve at the origin. Additionally, $X_3 = 1/3\cos(3\theta)$ introduces a third harmonic perturbation that intricately folds the curve three times along its path, creating a unique and complex shape that oscillates in both dimensions while incorporating the effects of the harmonic perturbation.

@@

Building on simple circular structures, the `gen_unifsphere(n, p, r)` function function extends the idea to three dimensions by generating $n$ observations approximately uniformly distributed on the surface of a sphere of radius $r$, with optional high-dimensional noise when $p > 3$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) c"} else { "Figure \\@ref(fig:sphere-proj) c"})`). Each observation is computed from spherical coordinates, with $u \sim U(-1, 1)$ representing $\cos(\phi)$ and $\theta \sim U(0, 2\pi)$ the azimuthal angle. Cartesian coordinates are then defined as $X_1 = r\sqrt{1 - u^2}\cos(\theta)$, $X_2 = r\sqrt{1 - u^2}\sin(\theta)$, and $X_3 = ru$, ensuring uniform distribution on the surface (not within) of the sphere. For $p > 3$, additional dimensions $X_4$ through $X_p$ are appended as structured noise.

@@

In addition, the `gen_gridedsphere(n, p)` function constructs a $p\text{-}D$ dataset consisting of approximately $n$ points that are evenly distributed on the surface of the unit $(p-1)$-sphere embedded in $\mathbb{R}^p$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) d"} else { "Figure \\@ref(fig:sphere-proj) d"})`). The method relies on forming a regular grid in spherical coordinates, parameterized by $(p-1)$ angular variables: for dimensions $j = 1, \dots, p-2$ the polar angles are drawn from $[0, \pi]$, while the final angle ($j = p-1$) represents the azimuth and is drawn from $[0, 2\pi]$. The number of grid steps along each angular dimension is chosen by decomposing $n$ into $(p-1)$ approximately equal integer factors using the helper function `gen_nproduct(n, p - 1)`.

Each grid point is subsequently mapped into Cartesian space via the standard hyperspherical-to-Cartesian transformation,

@@

The result is a deterministic grid of points lying exactly on the surface of the unit $(p-1)$-sphere, without any additional noise dimensions.

@@

For more heterogeneous structures, the `gen_clusteredspheres(n, k, p, r, loc)` function generates one large sphere of radius $r_1$ and $k$ smaller spheres of radius $r_2$, each centered at a different random location (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) e"} else { "Figure \\@ref(fig:sphere-proj) e"})`). A large uniform sphere centered at the origin is created by sampling $n_1$ points uniformly on the surface of a $p\text{-}D$ sphere with a radius of $r_1$. The sampling is executed using the function `gen_unifsphere(n_1, p, r_1)`, which generates the desired points in the specified dimensional space. In generation of $k$ smaller uniform spheres, each sphere contains $n_2$ points that are sampled uniformly on a sphere with a radius of $r_2$. These spheres are positioned at distinct random locations in $p$-space, with the center of each sphere being drawn from a normal distribution $N(0, \texttt{loc}^2 I_p)$. Points on spheres are generated using the standard hyperspherical method, which involves sampling $u \sim U(-1, 1)$ to determine the cosine of the polar angle, and sampling $\theta \sim U(0, 2\pi)$ to determine the azimuthal angle (for $3\text{-}D$). Each observation is classified by cluster, with labels such as "big" for the large central sphere and "small_1" to "small_k" for the smaller spheres.

@@

Finally, the `gen_hemisphere(n, p)` function restricts sampling to a hemisphere of a $4\text{-}D$ sphere (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) f"} else { "Figure \\@ref(fig:sphere-proj) f"})`). Using spherical coordinates, the azimuthal angle $\theta_1 \sim U(0, \pi)$ in the $(x_1, x_2)$ plane, while the elevation angle $\theta_2 \sim U(0, \pi)$ in the $(x_2, x_3)$ plane. Additionally, $\theta_3 \sim U(0, \pi/2)$ in the $(x_3, x_4)$ plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into $4\text{-}D$ Cartesian space: $X_1 = \sin(\theta_1)\cos(\theta_2)$, $X_2 = \sin(\theta_1)\sin(\theta_2)$, $X_3 = \cos(\theta_1)\cos(\theta_3)$, $X_4 = \cos(\theta_1)\sin(\theta_3)$. This produces points on one side of a $4\text{-}D$ unit sphere, effectively generating a $4\text{-}D$ hemisphere. For $p > 4$, additional noise dimensions are added to embed the structure in higher dimensions.

### 2025-09-20 paper-cardinalR.tex
Updated sphere section

\subsubsection{Sphere}\label{sphere}

Sphere-shaped structures (Figure \ref{fig:sphere-proj}) are useful for evaluating how dimension reduction and clustering algorithms handle curved, symmetric manifolds in high-dimensional spaces. The package provides functions to generate a variety of spherical forms, including simple circles, uniform spheres, grid-based spheres, and complex arrangements like clustered spheres within a larger sphere. Table \ref{tab:sphere-tb-pdf} summarizes these functions.

@@

\paragraph{\texorpdfstring{\texttt{gen\_circle()}}{gen\_circle()}}\label{gen_circle}

The function \texttt{gen\_circle(n,\ p)} generates a \(p\)-dimensional dataset of \(n\) observations, where the first two dimensions form a unit circle, and the remaining dimensions are structured sinusoidal extensions of the angular parameter with progressively smaller scale (Figure \ref{fig:sphere-proj} a).

A latent angle variable \(\theta\) is uniformly sampled from the interval \([0, 2\pi]\). Coordinates in the first two dimensions represent a perfect circle on the plane: \(X_1 = \cos(\theta), \quad X_2 = \sin(\theta)\). For dimensions \(X_3\) through \(X_p\), sinusoidal transformations of the angle \(\theta\) are introduced. The first component is a scaling factor that decreases with the dimension index, defined as \(\text{scale}_j = \sqrt{(0.5)^{j-2}}\) for \(j = 3, \dots, p\). The second component is a phase shift that is proportional to the dimension index, specifically designed to decorrelate the curves, given by the formula \(\phi_j = (j - 2)\pi/2p\). Each additional dimension is computed as: \(X_j = \text{scale}_{j}\sin(\theta + \phi_j), \quad j = 3, \dots, p\).

@@

\paragraph{\texorpdfstring{\texttt{gen\_curvycycle()}}{gen\_curvycycle()}}\label{gen_curvycycle}

The \texttt{gen\_curvycycle(n,\ p)} function generates a \(p\)-dimensional dataset of \(n\) observations lying on a curved closed loop with controlled high-dimensional sinusoidal deviations (Figure \ref{fig:sphere-proj} b). The structure forms a one-dimensional nonlinear cycle embedded in higher-dimensional space.

A latent angle variable \(\theta\) is uniformly sampled from the interval \([0, 2\pi]\). The first three dimensions define a non-circular closed curve, referred to as a ``curvy cycle''. In this configuration, \(X_1 = \cos(\theta)\) represents horizontal oscillation, while \(X_2 = \sqrt{3}/3 + \sin(\theta)\) introduces a vertical offset to avoid centering the curve at the origin. Additionally, \(X_3 = 1/3\cos(3\theta)\) introduces a third harmonic perturbation that intricately folds the curve three times along its path, creating a unique and complex shape that oscillates in both dimensions while incorporating the effects of the harmonic perturbation.

@@

\paragraph{\texorpdfstring{\texttt{gen\_unifsphere()}}{gen\_unifsphere()}}\label{gen_unifsphere}

The \texttt{gen\_unifsphere(n,\ p,\ r)} function generates a \(p\)-dimensional dataset of \(n\) observations distributed approximately uniformly on the surface of a \(3\text{-}D\) sphere of radius \(r\), with additional Gaussian noise dimensions added when \(p > 3\) (Figure \ref{fig:sphere-proj} c).

Each observation lies on the surface of a sphere in \(3\text{-}D\), constructed by generating \(u \sim U(-1, 1)\) which represents the cosine of the polar angle \(\phi\) and \(\theta \sim U(0, 2\pi)\) which represents the azimuthal angle. The corresponding Cartesian coordinates are calculated as; \(X_1 = r\sqrt{1 - u^2}\cos(\theta)\), \(X_2 = r\sqrt{1 - u^2}\sin(\theta)\), \(X_3 = ru\), which gives points uniformly distributed on the surface of a \(3\text{-}D\) sphere (not within). For \(p > 3\), additional noise dimensions \(X_4\) to \(X_p\) are generated.

@@

\paragraph{\texorpdfstring{\texttt{gen\_gridedsphere()}}{gen\_gridedsphere()}}\label{gen_gridedsphere}

The \texttt{gen\_gridedsphere(n,\ p)} function generates a \(p\)-dimensional dataset of approximately \(n\) points evenly distributed on the surface of a \((p!-!1)\)-sphere (the unit sphere embedded in \(\mathbb{R}^p\)) (Figure \ref{fig:sphere-proj} d).

The construction is based on a regular grid in spherical coordinates with \((p-1)\) angular variables. Specifically, for dimensions \(j = 1, \dots, p-2\), the polar angles are defined on \([0, \pi]\). For the last angle (\(j = p-1\)), the azimuthal angle is defined on \([0, 2\pi]\).

The number of grid steps along each angular dimension is determined by factoring \(n\) into \((p-1)\) approximately equal integers via \texttt{gen\_nproduct(n,\ p\ -\ 1)}.

Each grid point is then mapped to Cartesian coordinates using the standard hyperspherical-to-Cartesian transformation:

@@

Thus, the output is a deterministic grid of points lying exactly on the unit \((p-1)\)-sphere, with no added noise dimensions.

@@

\paragraph{\texorpdfstring{\texttt{gen\_clusteredspheres()}}{gen\_clusteredspheres()}}\label{gen_clusteredspheres}

The \texttt{gen\_clusteredspheres(n,\ k,\ p,\ r,\ loc)} function generates a synthetic dataset of \(n_1 + kn_2\) observations in \(p\)-dimensional space, consisting of one large sphere of radius \(r_1\) and \(k\) smaller spheres of radius \(r_2\), each centered at a different random location (Figure \ref{fig:sphere-proj} e).

A large uniform sphere centered at the origin is created by sampling \(n_1\) points uniformly on the surface of a \(p\text{-}D\) sphere with a radius of \(r_1\). The sampling is executed using the function \texttt{gen\_unifsphere(n\_1,\ p,\ r\_1)}, which generates the desired points in the specified dimensional space.

In generation of \(k\) smaller uniform spheres, each sphere contains \(n_2\) points that are sampled uniformly on a sphere with a radius of \(r_2\). These spheres are positioned at distinct random locations in \(p\)-space, with the center of each sphere being drawn from a normal distribution \(N(0, \texttt{loc}^2 I_p)\).

Points on spheres are generated using the standard hyperspherical method, which involves sampling \(u \sim U(-1, 1)\) to determine the cosine of the polar angle, and sampling \(\theta \sim U(0, 2\pi)\) to determine the azimuthal angle (for \(3\text{-}D\)). Each observation is classified by cluster, with labels such as ``big'' for the large central sphere and ``small\_1'' to ``small\_k'' for the smaller spheres.

@@

\paragraph{\texorpdfstring{\texttt{gen\_hemisphere()}}{gen\_hemisphere()}}\label{gen_hemisphere}

The \texttt{gen\_hemisphere(n,\ p)} function generates a \(p\)-dimensional dataset of \(n\) observations distributed approximately uniformly on a \(4\text{-}D\) hemisphere, optionally extended with Gaussian noise in additional dimensions when \(p > 4\) (Figure \ref{fig:sphere-proj} f).

Each observation is situated on a restricted \(4\text{-}D\) spherical surface, defined by spherical coordinates. The azimuthal angle \(\theta_1 \sim U(0, \pi)\) in the \((x_1, x_2)\) plane, while the elevation angle \(\theta_2 \sim U(0, \pi)\) in the \((x_2, x_3)\) plane. Additionally, \(\theta_3 \sim U(0, \pi/2)\) in the \((x_3, x_4)\) plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into \(4\text{-}D\) Cartesian space: \(X_1 = \sin(\theta_1)\cos(\theta_2)\), \(X_2 = \sin(\theta_1)\sin(\theta_2)\), \(X_3 = \cos(\theta_1)\cos(\theta_3)\), \(X_4 = \cos(\theta_1)\sin(\theta_3)\). This produces points on one side of a \(4\text{-}D\) unit sphere, effectively generating a \(4\text{-}D\) hemisphere. For \(p > 4\), additional noise dimensions are added to embed the structure in higher dimensions.

@@

\subsubsection{Trigonometric}\label{trigonometric}

@@

\subsubsection{Trefoil knots}\label{trefoil-knots}

>>>>

\paragraph{Sphere}\label{sphere}

Sphere-shaped structures (Figure \ref{fig:sphere-proj}) are useful for evaluating how dimension reduction and clustering algorithms handle curved, symmetric manifolds in high-dimensional spaces. The functions generate a variety of spherical forms, including simple circles, uniform spheres, grid-based spheres, and complex arrangements like clustered spheres within a larger sphere. The first few coordinates define the main geometric form (circle, cycle, sphere, or hemisphere), while higher-dimensional embeddings are achieved by adding noise dimensions. For the circle and curvy cycle, this occurs when \(p > 2\) and \(p > 3\), respectively, while for spheres and hemispheres it occurs when \(p > 3\) or \(p > 4\). Table \ref{tab:sphere-tb-pdf} summarizes these functions.

@@

The simplest case, \texttt{gen\_circle(n,\ p)} creates a unit circle in two dimensions, with the remaining dimensions forming sinusoidal extensions of the angular parameter at progressively smaller scales (Figure \ref{fig:sphere-proj} a). Let a latent angle variable \(\theta\) is uniformly sampled from the interval \([0, 2\pi]\). Coordinates in the first two dimensions represent a perfect circle on the plane: \(X_1 = \cos(\theta), \quad X_2 = \sin(\theta)\). For dimensions \(X_3\) through \(X_p\), sinusoidal transformations of the angle \(\theta\) are introduced. The first component is a scaling factor that decreases with the dimension index, defined as \(\text{scale}_j = \sqrt{(0.5)^{j-2}}\) for \(j = 3, \dots, p\). The second component is a phase shift that is proportional to the dimension index, specifically designed to decorrelate the curves, given by the formula \(\phi_j = (j - 2)\pi/2p\). Each additional dimension is computed as: \(X_j = \text{scale}_{j}\sin(\theta + \phi_j), \quad j = 3, \dots, p\).

@@

For the one-dimensional nonlinear cycle embedded in \(p\text{-}D\) space, \texttt{gen\_curvycycle(n,\ p)} (Figure \ref{fig:sphere-proj} b), let a latent angle variable \(\theta\) is uniformly sampled from the interval \([0, 2\pi]\). The first three dimensions define a non-circular closed curve, referred to as a ``curvy cycle''. In this configuration, \(X_1 = \cos(\theta)\) represents horizontal oscillation, while \(X_2 = \sqrt{3}/3 + \sin(\theta)\) introduces a vertical offset to avoid centering the curve at the origin. Additionally, \(X_3 = 1/3\cos(3\theta)\) introduces a third harmonic perturbation that intricately folds the curve three times along its path, creating a unique and complex shape that oscillates in both dimensions while incorporating the effects of the harmonic perturbation.

@@

Building on simple circular structures, the \texttt{gen\_unifsphere(n,\ p,\ r)} function function extends the idea to three dimensions by generating \(n\) observations approximately uniformly distributed on the surface of a sphere of radius \(r\), with optional high-dimensional noise when \(p > 3\) (Figure \ref{fig:sphere-proj} c). Each observation is computed from spherical coordinates, with \(u \sim U(-1, 1)\) representing \(\cos(\phi)\) and \(\theta \sim U(0, 2\pi)\) the azimuthal angle. Cartesian coordinates are then defined as \(X_1 = r\sqrt{1 - u^2}\cos(\theta)\), \(X_2 = r\sqrt{1 - u^2}\sin(\theta)\), and \(X_3 = ru\), ensuring uniform distribution on the surface (not within) of the sphere. For \(p > 3\), additional dimensions \(X_4\) through \(X_p\) are appended as structured noise.

@@

In addition, the \texttt{gen\_gridedsphere(n,\ p)} function constructs a \(p\text{-}D\) dataset consisting of approximately \(n\) points that are evenly distributed on the surface of the unit \((p-1)\)-sphere embedded in \(\mathbb{R}^p\) (Figure \ref{fig:sphere-proj} d). The method relies on forming a regular grid in spherical coordinates, parameterized by \((p-1)\) angular variables: for dimensions \(j = 1, \dots, p-2\) the polar angles are drawn from \([0, \pi]\), while the final angle (\(j = p-1\)) represents the azimuth and is drawn from \([0, 2\pi]\). The number of grid steps along each angular dimension is chosen by decomposing \(n\) into \((p-1)\) approximately equal integer factors using the helper function \texttt{gen\_nproduct(n,\ p\ -\ 1)}.

Each grid point is subsequently mapped into Cartesian space via the standard hyperspherical-to-Cartesian transformation,

@@

The result is a deterministic grid of points lying exactly on the surface of the unit \((p-1)\)-sphere, without any additional noise dimensions.

@@

For more heterogeneous structures, the \texttt{gen\_clusteredspheres(n,\ k,\ p,\ r,\ loc)} function generates one large sphere of radius \(r_1\) and \(k\) smaller spheres of radius \(r_2\), each centered at a different random location (Figure \ref{fig:sphere-proj} e). A large uniform sphere centered at the origin is created by sampling \(n_1\) points uniformly on the surface of a \(p\text{-}D\) sphere with a radius of \(r_1\). The sampling is executed using the function \texttt{gen\_unifsphere(n\_1,\ p,\ r\_1)}, which generates the desired points in the specified dimensional space. In generation of \(k\) smaller uniform spheres, each sphere contains \(n_2\) points that are sampled uniformly on a sphere with a radius of \(r_2\). These spheres are positioned at distinct random locations in \(p\)-space, with the center of each sphere being drawn from a normal distribution \(N(0, \texttt{loc}^2 I_p)\). Points on spheres are generated using the standard hyperspherical method, which involves sampling \(u \sim U(-1, 1)\) to determine the cosine of the polar angle, and sampling \(\theta \sim U(0, 2\pi)\) to determine the azimuthal angle (for \(3\text{-}D\)). Each observation is classified by cluster, with labels such as ``big'' for the large central sphere and ``small\_1'' to ``small\_k'' for the smaller spheres.

@@

Finally, the \texttt{gen\_hemisphere(n,\ p)} function restricts sampling to a hemisphere of a \(4\text{-}D\) sphere (Figure \ref{fig:sphere-proj} f). Using spherical coordinates, the azimuthal angle \(\theta_1 \sim U(0, \pi)\) in the \((x_1, x_2)\) plane, while the elevation angle \(\theta_2 \sim U(0, \pi)\) in the \((x_2, x_3)\) plane. Additionally, \(\theta_3 \sim U(0, \pi/2)\) in the \((x_3, x_4)\) plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into \(4\text{-}D\) Cartesian space: \(X_1 = \sin(\theta_1)\cos(\theta_2)\), \(X_2 = \sin(\theta_1)\sin(\theta_2)\), \(X_3 = \cos(\theta_1)\cos(\theta_3)\), \(X_4 = \cos(\theta_1)\sin(\theta_3)\). This produces points on one side of a \(4\text{-}D\) unit sphere, effectively generating a \(4\text{-}D\) hemisphere. For \(p > 4\), additional noise dimensions are added to embed the structure in higher dimensions.

@@

\paragraph{Trigonometric}\label{trigonometric}

@@

\paragraph{Trefoil knots}\label{trefoil-knots}
